Unlocking Sensitivity for Visibility-based Power Spectrum Estimators of the 21\,cm Reionization Signal

### 2017-07-18 Rotation paper/draft_working.tex
saving progress

Analysis pipelines for the 21\,cm power spectrum typically fall into two categories. In the first, images are formed in Fourier domain through rotation synthesis, followed by a foreground mitigation step to construct a power spectrum. An alternative technique works directly with visibilities from baselines, delay-transforming and cross-multiplying them to form the power spectrum. This technique avoids many systematics associated with combining data from different baselines and tracks the native sampling of the interferometer. An example of the visibility-based pipeline was presented in \cite{Ali2015}, which provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth as PAPER-64). However, one disadvantage of existing visibility-based pipelines is their lack of use of partial redundancy. Baselines of different lengths and orientations contain partially redundant information. While imaging based power-spectrum pipelines naturally include all redundancy information, visibility-based pipelines so far only cross multiply fully redundant baselines, i.e. baselines of the same length and orientation.

@@

Roughly speaking, Eqs. \eqref{eq:opp} through \eqref{eq:Thetanu} tell us that the cross multiplications of visibilities at a time delay is proportional to the power spectrum times the Fourier
transform of the cross multiplied beam patterns. As a check, when applied to equivalent baselines,
$\boldsymbol{b}=\boldsymbol{b'}$, the peak correlation occurs at $\Delta T=0$ and $\hat{\boldsymbol{s}}=\boldsymbol{\Gamma}\hat{\boldsymbol{s}}$, in which case Eq. \eqref{eq:final} reduces to Eq. (B9) of \cite{paper32}. 
With Eq. \eqref{eq:opp} and Eq. \eqref{eq:Theta} we can, for any given pair of baseline classes and time delay, estimate the degree of redundancy, here represented by $\Theta$. This allows us to achieve our goals stated in the introduction: to identify 
candidate baseline pairs with significant redundancy, to find the time offset that maximizes redundancy, and to quantify the degree of such redundancy. We can do all the above simply by computing the weight $\Theta$ from
Eq. \eqref{eq:Theta} for various time offsets.

@@

At the time offset of maximum redundancy, we expect two beams to have to same fringe pattern in both frequency and phase. Due to the time delay, however, the beam centers would be slightly shifted with respect to each other. This we show in the top panels of Fig. \ref{fig:beamfringe}. The top left and middle panels show the beam-fringe
patterns (real parts) for baselines \{2,0\} and \{2,1\}, delayed by 0.0325 sidereal days.
The product of the beam fringe patterns in top right shows that the fringes
 cancel out as we expect. On the bottom of Fig. \ref{fig:beamfringe} we show the instantaneous $uv$ coverage of two baselines and their product. We see from the bottom right panel that the cross-multiplied baselines recover  power concentrated at $u\approx\frac{30\text{m}}{c/1.5\times10^8\text{Hz}}\approx15$. Note the bottom right panel is not the Fourier transform of the top right panel.

@@

Figure \ref{fig:multiblwindows} shows four example window functions. The two window functions in the top plot are for equivalent baseline power spectrum estimators, i.e., where one squares the delay spectrum from a single baseline and normalizes the result. The window function localized at low $\ell$ is for the baseline \{2,0\} of around $30\,\textrm{m}$ in length, while the window function localized at high $\ell$ is for \{12,0\}, roughly $180\,\textrm{m}$ in length. The window functions for near-equivalent baseline estimates are shown in the bottom plot, for \{2,0:2,1\} and \{12,0:12,1\} at low- and high-$\bar{b}$, respectively. In all cases, the window functions that we show are for $\tau = 703\,\textrm{ns}$, and were generated using primary beam models for PAPER. For numerical convenience we assumed that the bandpass $\phi_\nu$ (see Eq. \eqref{eq:B}) takes the form of a half-period sine curve that peaks at $150\,\textrm{\,MHz}$ and goes to zero $5\,\textrm{\,MHz}$ on either side of the peak.

@@

\subsection{ Sensitivity \label{sec:sensitivity}}
In this section we discuss the sensitivity contributions of a variety of baseline pairs from PAPER-128.
Recall the sensitivity contribution of a particular baseline pair is given by $\Theta$ (Eq. \eqref{eq:Theta}). From now one we use $\Theta_{\boldsymbol{bb'}}$ and $\Delta T_{\boldsymbol{bb'}}$ to denote the peak value and location of $\Theta$, or in other words the maximum of the redundancy weight $|\Theta(\Delta T)|$ and the corresponding offset for a given baseline pair $\{\boldsymbol{b}:\boldsymbol{b'}\}$. 
%Intuitively we expect the sensitivity to depend on both the $uv$ coverage of the baseline and the patch of sky inside the beam. A larger beam like that of PAPER would tolerate larger time offsets because more sky area can coincide in the two beams \footnote{Larger beams also imply smaller spread in $uv$ space. This could lead to either larger or smaller redundancy, which depends on the overlap of two such point spread functions.}

In the middle panel of Fig. \ref{fig:sensplot}
we show $\Theta_{\boldsymbol{bb'}}$ and $\Delta T_{\boldsymbol{bb'}}$ for a variety of baseline combinations. Baseline pairs that are mirror images of each other 
give the same amount of redundancies ($\Theta_{\boldsymbol{bb'}}$), with the opposite time offset, as expected from symmetry. 
For example, \{1,0:1,1\} is mirror image of \{1,0:1,-1\} and these two pairs of baselines
give the same sensibility contribution. Thus we only show a subset of representative baseline pairs to illustrate the contributions. For a more complete result, see Fig. \ref{fig:pairplot}.

@@

The scaling in Eq. \eqref{eq:sensul} was rough for simplicity of motivation. As we derive in Appendix \ref{sec:appB}, this weight should be corrected by a factor proportional $\rho_0$:

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time delay $\Delta T_{\boldsymbol{bb'}}$,  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$ and baseline length $\bar{b}$. Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Only one baseline length is shown since all top contributing pairs have very similar lengths as expected. The scatter plots are shown with transparency so that darker regions indicate degeneracies. Scatter points of the 3 arrays overlap, in order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

In Fig. \ref{fig:pairplot} we present pair-distributions of three different properties for baselines that contribute well to the sensitivity ($\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$, where again the weight for the top class of equivalent pair is normalized to 1). The properties shown are effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, optimal time offset $\Delta T_{\boldsymbol{bb'}}$, and average baseline length $\bar{b}$. We only show three of the mentioned arrays for visual clarity. The other two results are qualitatively similar. We point out some key features:
\begin{itemize}

\item $\Delta T_{\boldsymbol{bb'}}$ vs. $\widetilde{\Theta}_{\boldsymbol{bb'}}$:
This relation  is familiar from Fig. \ref{fig:sensplot}. The points at $\Delta T=0$ are the equivalent baselines. Note HERA arrays have fewer data points with high $\Delta T_{\boldsymbol{bb'}}$, due to their comparatively narrow beam.

\item $\Delta T_{\boldsymbol{bb'}}$ vs. $\bar{b}$:
PAPER-128 shows the trend that longer baselines correspond to lower $\Delta T_{\boldsymbol{bb'}}$. HERA arrays do not exhibit this trend here because of a selection effect. Only baseline pairs with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown. Shorter HERA baselines require much longer $\Delta T$ to overlap, partly because of the hexagonal structure requires $60\degree$ of rotation to overlap baselines, partly because of the smaller primary beam. Thus most short HERA baseline pairs are not shown in this figure.

@@

Current visibility-based power spectrum pipelines have the advantage of easy-to-manage systematics, but have yet to unlock the full sensitivity of the arrays as they do not incorporate partially redundant baselines.  We present a visibility and delay-transform based method to overcome this limitation by extracting the sensitivity contained in such partial redundancy. Applicable to both tracking and drift scan measurements, our method relies on cross-multiplications of visibilities that are offset in time. Given an antenna array configuration, our method identifies the best baseline pairs to cross-multiply and predicts the optimal time offset $\Delta T_{\boldsymbol{bb'}}$, weight $\Theta_{\boldsymbol{bb'}}$, and the necessary rephasing factor $\psi_\nu$. With the predicted results one can incorporate partial redundancy into existing delay-transform based power-spectrum pipelines.

@@

Track-crossing corresponds to vanishing of the exponent
\[
 \boldsymbol{b}-\boldsymbol{\Gamma}^{-1}\boldsymbol{b'} = 0. \label{eq:trackcross}
 \]
If exponent vanishes, $\Theta_{\nu}(\Delta T)$ is real, and one would not observe any non-zero phase at its peak value. However, with the inclusion of the $w$-term, we see that track crossings in $uv$-plane do not actually imply crossing in the $uvw$ space, and therefore the exponent or, equivalently, Eq. \eqref{eq:trackcross} does not vanish and we observe non-zero, frequency dependent peak phases even in the case of tracking arrays, as shown in Fig. \ref{fig:phi_nu}.

>>>>

Analysis pipelines for the 21\,cm power spectrum typically fall into two categories. In the first, images are formed in Fourier domain through rotation synthesis, followed by a foreground mitigation step to construct a power spectrum. An alternative technique works directly with visibilities from baselines, delay-transforming and cross-multiplying them to form the power spectrum. This technique avoids many systematics associated with combining data from different baselines and tracks the native sampling of the interferometer. An example of the visibility-based pipeline is presented in \cite{Ali2015}, which provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth as PAPER-64). However, one disadvantage of existing visibility-based pipelines is their lack of use of partial redundancy. Baselines of different lengths and orientations contain partially redundant information. While imaging based power-spectrum pipelines naturally include all redundancy information, visibility-based pipelines so far only cross multiply fully redundant baselines, i.e. baselines of the same length and orientation.

@@

Roughly speaking, Eqs. \eqref{eq:opp} through \eqref{eq:Thetanu} tell us that the cross multiplications of visibilities at a time delay is proportional to the power spectrum times the Fourier
transform of the cross multiplied beam patterns. As a check, when applied to equivalent baselines,
$\boldsymbol{b}=\boldsymbol{b'}$, the peak correlation occurs at $\Delta t=0$ and $\hat{\boldsymbol{s}}=\boldsymbol{\Gamma}\hat{\boldsymbol{s}}$, in which case Eq. \eqref{eq:final} reduces to Eq. (B9) of \cite{paper32}. 
With Eq. \eqref{eq:opp} and Eq. \eqref{eq:Theta} we can, for any given pair of baseline classes and time delay, estimate the degree of redundancy, here represented by $\Theta$. This allows us to achieve our goals stated in the introduction: to identify 
candidate baseline pairs with significant redundancy, to find the time offset that maximizes redundancy, and to quantify the degree of such redundancy. We can do all the above simply by computing the weight $\Theta$ from
Eq. \eqref{eq:Theta} for various time offsets.

@@

At the time offset of maximum redundancy, we expect two baselines to have to same fringe pattern in both frequency and phase. Due to the time delay, however, the beam centers would be slightly shifted with respect to each other. In the top panels of Fig. \ref{fig:beamfringe}, we show the real parts of peak-normalized beam-fringe patterns. The top left and middle panels show the beam-fringe
patterns for the baseline classes \{2,0\} and \{2,1\}, offset by 0.0325 sidereal days. More specifically, displayed are the real parts of $B(\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\hat{\boldsymbol{s}}\cdot\boldsymbol{b}\right]$ and $B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right]$, respectively. The top right panel show their conjugate product $B^{*}(\hat{\boldsymbol{s}},\nu)B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[i2\pi\frac{\nu}{c}\left(\hat{\boldsymbol{s}}\cdot\boldsymbol{b}-\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right)-i\psi_{\nu}\right]$, where we have chosen global phase $\psi_{\nu}$ such that the peak of the conjugate product is real. The product of the beam fringe patterns in top right shows that the fringes
 cancel out as we expect. Comparing with Eq. \eqref{eq:Thetanu}, we see that $\Theta_\nu$ is the sum over the top right plot. Thus we show that at the optimal time offset the fringe patterns of the two baselines cancel out, leading to a substantial weight $\Theta_\nu$. The frequency-summed weight $\Theta$ is constructed with rephasing as described in Section \ref{sec:rephs} so that the conjugate product (top right panel) for different frequencies align in phase.

Similarly, we show the correlation of the two baselines in terms of their $uv$ beam spread overlap. The bottom left and middle panels are (peak-normalized) instantaneous $uv$ coverage of two baselines while the bottom right panel is their product. We see from the bottom right panel that the cross-multiplied baselines recover power concentrated at $u\approx15$. The product of the $uv$ beams spread shown in the bottom right panel should not be confused with the Fourier transform of the beam-fringe conjugate product shown in the top right panel.

@@

Figure \ref{fig:multiblwindows} shows four example window functions. The two window functions in the top plot are for equivalent baseline power spectrum estimators, i.e., where one squares the delay spectrum from a single baseline and normalizes the result. The window function localized at low $\ell$ is for the baseline \{2,0\} of around $30\,\textrm{m}$ in length, while the window function localized at high $\ell$ is for \{12,0\}, roughly $180\,\textrm{m}$ in length. The window functions for near-equivalent baseline estimates are shown in the bottom plot, for \{2,0:2,1\} and \{12,0:12,1\} at low- and high-$\bar{b}$, respectively. In all cases, the window functions that we show are for $\tau = 703\,\textrm{ns}$, and are generated using primary beam models for PAPER. For numerical convenience we assumed that the bandpass $\phi_\nu$ (see Eq. \eqref{eq:B}) takes the form of a half-period sine curve that peaks at $150\,\textrm{\,MHz}$ and goes to zero $5\,\textrm{\,MHz}$ on either side of the peak.

@@

\subsection{ Sensitivity \label{sec:sensitivity}}
In this section we discuss the sensitivity contributions of a variety of baseline pairs from PAPER-128.
Recall the sensitivity contribution of a particular baseline pair is given by $\Theta$ (Eq. \eqref{eq:Theta}). From now one we use $\Theta_{\boldsymbol{bb'}}$ and $\Delta t_{\boldsymbol{bb'}}$ to denote the peak value and location of $\Theta$, or in other words the maximum of the redundancy weight $|\Theta(\Delta t)|$ and the corresponding offset for a given baseline pair $\{\boldsymbol{b}:\boldsymbol{b'}\}$. 
%Intuitively we expect the sensitivity to depend on both the $uv$ coverage of the baseline and the patch of sky inside the beam. A larger beam like that of PAPER would tolerate larger time offsets because more sky area can coincide in the two beams \footnote{Larger beams also imply smaller spread in $uv$ space. This could lead to either larger or smaller redundancy, which depends on the overlap of two such point spread functions.}

In the middle panel of Fig. \ref{fig:sensplot}
we show $\Theta_{\boldsymbol{bb'}}$ and $\Delta t_{\boldsymbol{bb'}}$ for a variety of baseline combinations. Baseline pairs that are mirror images of each other 
give the same amount of redundancies ($\Theta_{\boldsymbol{bb'}}$), with the opposite time offset, as expected from symmetry. 
For example, \{1,0:1,1\} is mirror image of \{1,0:1,-1\} and these two pairs of baselines
give the same sensibility contribution. Thus we only show a subset of representative baseline pairs to illustrate the contributions. For a more complete result, see Fig. \ref{fig:pairplot}.

@@

The scaling in Eq. \eqref{eq:sensul} is rough for simplicity of motivation. As we derive in Appendix \ref{sec:appB}, this weight should be corrected by a factor proportional $\rho_0$:

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time delay $\Delta t_{\boldsymbol{bb'}}$,  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$ and baseline length $\bar{b}$. Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Only one baseline length is shown since all top contributing pairs have very similar lengths as expected. The scatter plots are shown with transparency so that darker regions indicate degeneracies. Scatter points of the 3 arrays overlap, in order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

In Fig. \ref{fig:pairplot} we present pair-distributions of three different properties for baselines that contribute well to the sensitivity ($\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$, where again the weight for the top class of equivalent pair is normalized to 1). The properties shown are effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, optimal time offset $\Delta t_{\boldsymbol{bb'}}$, and average baseline length $\bar{b}$. We only show three of the mentioned arrays for visual clarity. The other two results are qualitatively similar. We point out some key features:
\begin{itemize}

\item $\Delta t_{\boldsymbol{bb'}}$ vs. $\widetilde{\Theta}_{\boldsymbol{bb'}}$:
This relation  is familiar from Fig. \ref{fig:sensplot}. The points at $\Delta t=0$ are the equivalent baselines. Note HERA arrays have fewer data points with high $\Delta t_{\boldsymbol{bb'}}$, due to their comparatively narrow beam.

\item $\Delta t_{\boldsymbol{bb'}}$ vs. $\bar{b}$:
PAPER-128 shows the trend that longer baselines correspond to lower $\Delta t_{\boldsymbol{bb'}}$. HERA arrays do not exhibit this trend here because of a selection effect. Only baseline pairs with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown. Shorter HERA baselines require much longer $\Delta t$ to overlap, partly because of the hexagonal structure requires $60\degree$ of rotation to overlap baselines, partly because of the smaller primary beam. Thus most short HERA baseline pairs are not shown in this figure.

@@

Current visibility-based power spectrum pipelines have the advantage of easy-to-manage systematics, but have yet to unlock the full sensitivity of the arrays as they do not incorporate partially redundant baselines.  We present a visibility and delay-transform based method to overcome this limitation by extracting the sensitivity contained in such partial redundancy. Applicable to both tracking and drift scan measurements, our method relies on cross-multiplications of visibilities that are offset in time. Given an antenna array configuration, our method identifies the best baseline pairs to cross-multiply and predicts the optimal time offset $\Delta t_{\boldsymbol{bb'}}$, weight $\Theta_{\boldsymbol{bb'}}$, and the necessary rephasing factor $\psi_\nu$. With the predicted results one can incorporate partial redundancy into existing delay-transform based power-spectrum pipelines.

@@

Track-crossing corresponds to vanishing of the exponent
\[
 \boldsymbol{b}-\boldsymbol{\Gamma}^{-1}\boldsymbol{b'} = 0. \label{eq:trackcross}
 \]
If exponent vanishes, $\Theta_{\nu}(\Delta t)$ is real, and one would not observe any non-zero phase at its peak value. However, with the inclusion of the $w$-term, we see that track crossings in $uv$-plane do not actually imply crossing in the $uvw$ space, and therefore the exponent or, equivalently, Eq. \eqref{eq:trackcross} does not vanish and we observe non-zero, frequency dependent peak phases even in the case of tracking arrays, as shown in Fig. \ref{fig:phi_nu}.
